% Theoretical Analysis: Entropy Bounds Under Action Masking
% For inclusion in main paper or appendix

\section{Theoretical Analysis of Entropy Collapse}
\label{sec:entropy_theory}

We provide formal analysis of why entropy collapse is inevitable in RL with hierarchical action masking.

\subsection{Preliminaries}

Let $\mathcal{A}$ be the full action space with $|\mathcal{A}| = n$ actions. At state $s$, let $\mathcal{A}(s) \subseteq \mathcal{A}$ be the set of valid actions after masking, with $|\mathcal{A}(s)| = k(s) \leq n$.

For a policy $\pi$, the entropy at state $s$ is:
\begin{equation}
H(\pi(\cdot|s)) = -\sum_{a \in \mathcal{A}(s)} \pi(a|s) \log \pi(a|s)
\end{equation}

\subsection{Upper Bound on Entropy}

\begin{theorem}[Entropy Upper Bound Under Masking]
\label{thm:entropy_bound}
For any policy $\pi$ at state $s$ with $k(s)$ valid actions:
\begin{equation}
H(\pi(\cdot|s)) \leq \log k(s)
\end{equation}
with equality iff $\pi$ is uniform over $\mathcal{A}(s)$.
\end{theorem}

\begin{proof}
Direct application of maximum entropy principle. The uniform distribution maximizes entropy over a finite support.
\end{proof}

\begin{corollary}[Entropy Collapse Threshold]
When $k(s) \leq 2$, the maximum possible entropy is $\log 2 \approx 0.693$ nats. For deterministic situations where $k(s) = 1$, entropy is exactly zero.
\end{corollary}

\subsection{Expected Entropy Under Stochastic Masking}

In hierarchical scheduling, the number of valid actions depends on system state. Let $K$ be a random variable representing $k(s)$ under the state distribution $d^\pi$.

\begin{theorem}[Expected Entropy Bound]
\label{thm:expected_entropy}
The expected entropy under policy $\pi$ satisfies:
\begin{equation}
\mathbb{E}_{s \sim d^\pi}[H(\pi(\cdot|s))] \leq \mathbb{E}[K] \cdot H\left(\frac{1}{\mathbb{E}[K]}\right) = \log \mathbb{E}[K]
\end{equation}
by Jensen's inequality (log is concave).
\end{theorem}

\subsection{Dynamics of Valid Action Count}

In shipyard scheduling, we observe:

\begin{lemma}[Valid Action Bottleneck]
Let $n_v$ be the number of vehicles (SPMTs) and $n_c$ be the number of cranes. The number of valid actions $k(s)$ satisfies:
\begin{equation}
k(s) \leq \min(n_v, |\{b : \text{block } b \text{ ready for transport}\}|) + \min(n_c, |\{b : \text{block } b \text{ ready for erection}\}|) + 1
\end{equation}
where $+1$ accounts for the HOLD action.
\end{lemma}

In practice, when most vehicles are busy transporting blocks or most cranes are lifting, $k(s) \to 1$ (only HOLD is valid), forcing entropy to zero.

\subsection{Policy Gradient Under Low Entropy}

\begin{theorem}[Gradient Vanishing at Low Entropy]
\label{thm:gradient_vanishing}
For a softmax policy $\pi_\theta(a|s) \propto \exp(f_\theta(s,a))$, the policy gradient magnitude satisfies:
\begin{equation}
\|\nabla_\theta \log \pi_\theta(a|s)\| \leq \frac{1}{\min_{a'} \pi_\theta(a'|s)}
\end{equation}
When $\pi$ approaches deterministic (low entropy), gradients for non-dominant actions vanish exponentially.
\end{theorem}

\begin{proof}
For softmax, $\nabla_\theta \log \pi_\theta(a|s) = \nabla_\theta f_\theta(s,a) - \mathbb{E}_{\pi}[\nabla_\theta f_\theta(s,\cdot)]$. As $\pi \to \delta_a$ (Dirac at action $a$), the expectation concentrates, making gradients for alternative actions approach zero.
\end{proof}

\subsection{Implications for RL Training}

\begin{corollary}[Exploration Death Spiral]
Combining Theorems~\ref{thm:entropy_bound}, \ref{thm:expected_entropy}, and \ref{thm:gradient_vanishing}:
\begin{enumerate}
    \item Action masking constrains maximum entropy to $\log k(s)$
    \item Low $k(s)$ states force low entropy regardless of policy
    \item Low entropy leads to deterministic policies
    \item Deterministic policies have vanishing gradients for exploration
    \item Without exploration, policy cannot discover better actions
\end{enumerate}
This creates an irreversible ``death spiral'' towards deterministic, suboptimal policies.
\end{corollary}

\subsection{Why Imitation Learning Succeeds}

DAgger avoids this spiral because:
\begin{enumerate}
    \item \textbf{No exploration required}: Learning is supervised, not through trial-and-error
    \item \textbf{Dense expert signal}: Every state has a correct action label
    \item \textbf{Policy update via classification}: Cross-entropy loss has well-behaved gradients even for rare actions
\end{enumerate}

\begin{theorem}[DAgger Convergence Under Masking]
Under standard DAgger assumptions (bounded loss, realizability), the learned policy $\hat{\pi}$ satisfies:
\begin{equation}
J(\pi^*) - J(\hat{\pi}) \leq O\left(\frac{1}{N}\right) + \epsilon_{approx}
\end{equation}
where $N$ is the number of DAgger iterations and $\epsilon_{approx}$ is the function approximation error. Crucially, this bound is independent of entropy or exploration.
\end{theorem}

\subsection{Empirical Validation}

Table~\ref{tab:entropy_theory_validation} shows empirical confirmation across three domains:

\begin{table}[h]
\centering
\caption{Empirical validation of entropy bounds across domains}
\label{tab:entropy_theory_validation}
\begin{tabular}{lccccc}
\toprule
Domain & $|\mathcal{A}|$ & $\mathbb{E}[k(s)]$ & Predicted $H_{max}$ & Observed $H_{final}$ & Collapse? \\
\midrule
Shipyard (HHI) & 50+ & 2.3 & 0.83 & 0.00 & Yes \\
Job Shop (10×5) & 51 & 3.1 & 1.13 & 0.08 & Yes \\
VRPTW (20 cust.) & 21 & 4.2 & 1.44 & 0.12 & Yes \\
\bottomrule
\end{tabular}
\end{table}

In all domains, final entropy is far below the theoretical maximum, confirming entropy collapse.

\subsection{Mitigation Strategies and Their Limitations}

We analyze why common remedies fail:

\textbf{Entropy bonus}: Adding $\beta H(\pi)$ to the objective cannot increase entropy beyond $\log k(s)$.

\textbf{Softened masking}: Using finite penalties ($-M$ instead of $-\infty$) allows gradients but introduces invalid action probability, leading to constraint violations.

\textbf{Adaptive entropy tuning (SAC)}: Automatic temperature adjustment delays but does not prevent collapse when $k(s)$ remains small.

\textbf{Curriculum learning}: Training on larger $k(s)$ instances first helps but transfer fails when target domain has smaller $k(s)$.

\subsection{Conclusion}

Entropy collapse under action masking is a fundamental limitation of policy gradient methods, not a hyperparameter tuning issue. For hierarchical scheduling with constrained action spaces, imitation learning provides a principled solution that avoids the exploration-exploitation dilemma entirely.
